Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-22 paper.tex
Tidy up Agda, paper typos

Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_0$). Now, we can
eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
\Tm_0\,\Gamma\,\Bool_0$, we also have $\up\!b :
\Tm_1\,\Gamma\,(\Lift_1\,\Bool_0)$, hence $\up\!b :
\Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
$\mathsf{if}\,\up\!b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
\Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
computes appropriately on canonical terms: if $b$ is $\true_0$, we get
$\up\!\true_0 = \true_1$ as the if-then-else scrutinee.

@@

\textbf{Coercive cumulative subtyping.}
TTFL as specified does not directly support cumulative subtyping. However, it is
compatible with coercive subtyping. Consider the following rules:
\begin{alignat*}{3}
  &\blank\!\leq\!\blank &&: \Ty\,\Gamma\,i \to \Ty\,\Gamma\,j \to \Seti\\
  &\msf{coerce} &&: A \leq B \to \Tm\,\Gamma\,A \to \Tm\,\Gamma\,B\\
  &\leq\!\msf{refl} &&: A \leq A \\
  &\U\!\leq    &&: i < i' \to \U\,i\,j\,p \leq \U\,i'\,k\,q\\
  &\Pi\!\leq   &&: (p : A' \leq A) \to ((a' : \Tm\,\Gamma\,A')
  \to B[x \mapsto \msf{coerce}\,p\,a'] \leq B'[x \mapsto a'])\\
  & && \hspace{1em} \to \Pi(x : A) B \leq \Pi (x : A') B'
\end{alignat*}
Any model of TTFL can support the above rules: we can define
$\blank\!\leq\!\blank$ and $\msf{coerce}$ by indexed induction-recursion
\cite{indexedir}, where we define coercion along $\U\!\leq$ by type lifting, and
coercion along $\Pi\!\leq$ by backwards-forwards coercion.

@@

\textbf{Agda 2.6.1} has $\omega+1$-many non-cumulative predicative universes as
$\Set{i}$, with optional cumulative subtyping only for universes (and not other
type formers) \cite{agdadocs}.  It also has an internal type $\msf{Level} : \Set0$ for finite
levels (hence, exluding $\omega$), which supports constructors and some built-in
operations, but no general elimination rule. There is also an $\omega$-bounded
parallel hierarchy $\msf{Prop_i}$ for strict propositions \cite{sprop}. Agda
2.6.2 will extend the hierarchy to $\omega * 2$.

\textbf{Coq 8.13} has $\omega$-many cumulative predicative universes with
cumulative subtyping for all type formers \cite{timany18cumulative}. It supports
bounded universe polymorphism, but it has no internal type for levels, and
universe polymorphic definition are not internally typeable. It also optionally
has an impredicative $\msf{Prop}$ universe and optionally impredicative bottom
$\Seti$ universe. Version 8.13 added experimental support for a parallel
$\omega$-bounded cumulative hierarchy for strict propositions.

@@

\textbf{Idris 1} has countable cumulative universes with cumulative subtyping
only for universes, typical-ambiguity-style level inference and no universe
polymorphism \cite{idrisdocs}.

>>>>

Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_1$). Now, we can
eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
\Tm_0\,\Gamma\,\Bool_0$, we also have $\up\!b :
\Tm_1\,\Gamma\,(\Lift\,\Bool_0)$, hence $\up\!b :
\Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
$\mathsf{if}\,\up\!b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
\Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
computes appropriately on canonical terms: if $b$ is $\true_0$, we get
$\up\!\true_0 = \true_1$ as the if-then-else scrutinee.

@@

\textbf{Coercive cumulative subtyping.}
TTFL as specified does not directly support cumulative subtyping. However, it is
compatible with coercive subtyping. Consider the following rules:
\begin{alignat*}{3}
  &\blank\!\leq\!\blank &&: \Ty\,\Gamma\,i \to \Ty\,\Gamma\,j \to \Seti\\
  &\msf{coerce} &&: A \leq B \to \Tm\,\Gamma\,A \to \Tm\,\Gamma\,B\\
  &\leq\!\msf{refl} &&: A \leq A \\
  &\U\!\leq    &&: i < i' \to \U\,i\,j\,p \leq \U\,i'\,k\,q\\
  &\Pi\!\leq   &&: (p : A' \leq A) \to ((a' : \Tm\,\Gamma\,A')
  \to B[x \mapsto \msf{coerce}\,p\,a'] \leq B'[x \mapsto a'])\\
  & && \hspace{1em} \to \Pi(x : A) B \leq \Pi (x : A') B'
\end{alignat*}
Any model of TTFL can support the above rules: we can define
$\blank\!\leq\!\blank$ and $\msf{coerce}$ by indexed induction-recursion
\cite{indexedir}, where we define coercion along $\U\!\leq$ by type lifting, and
coercion along $\Pi\!\leq$ by backwards-forwards coercion. It is possible to
extend the subtyping relation with rules for other basic type formers.

@@

\textbf{Agda 2.6.1} has $\omega+1$-many non-cumulative predicative universes as
$\Set{i}$, with optional cumulative subtyping only for universes
\cite{agdadocs}.  It also has an internal type $\msf{Level} : \Set0$ for finite
levels (hence, exluding $\omega$), which supports constructors and some built-in
operations, but no general elimination rule. There is also a countable parallel
hierarchy $\msf{Prop_i}$ for strict propositions \cite{sprop}. Agda 2.6.2 will
extend the $\Set{i}$ hierarchy to $\omega * 2$.

\textbf{Coq 8.13} has $\omega$-many cumulative predicative universes with
cumulative subtyping for all type formers \cite{timany18cumulative}. It supports
bounded universe polymorphism, but it has no internal type for levels, and
universe polymorphic definitions are not internally typeable. It also has an
impredicative $\msf{Prop}$ universe and optionally impredicative bottom $\Seti$
universe. Version 8.13 added experimental support for a parallel
countable cumulative hierarchy for strict propositions.

@@

\textbf{Idris 1} has countable cumulative predicative universes with cumulative
subtyping only for universes, typical-ambiguity-style level inference and no
universe polymorphism \cite{idrisdocs}.
